Exploring probability distributions for bivariate temporal granularities

### 2019-11-18 main.Rmd
revision till smart meter

The grammar of graphics introduced a framework to incorporate underlying features for statistical graphics to construct them by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is a modification over the former, allows us to produce graphics using the same structured thinking that is used to design analysis. Briefly, any graphics are made up of **Layers**, **Facet**,**Scale** and a  **coordinate system**.**Layers** are made up of **data**, **mapping**, **statistical transformation**, **geometric objects** and **position adjustments**.

@@

There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or empirical statistical methods. The latter do not allow us to see the shape, skewness or nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. Density plots which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilties in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. The density based visualizations are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

@@

It is good to be conversant with the benefits and challenges while choosing any distribution plot, depending on the context. As a general rule, if we have too many categories, the quantile plots are useful for camparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior.

### Levels of temporal granularities

The levels of the two granularities plotted might have an impact on the choice of plots since space and resolution might become a problem if the number of levels are too high. The criteria for different levels could be based on usual cognitive power while comparing across facets and display size available to us. Plot choices will also vary depending on which granularity is placed on the x-axis and which one across facets.

@@

  - If level of the granularity plotted across x-axis is low or medium, then any distribution plots might be chosen depending on which feature of the distribution needs focus, irrespective of the number of the levels of the facet variable.

@@

  - - If levels of granularities plotted are medium, empirical based methods of statistical distribution are preferred as they use less space by design than most density based methods.

@@

## Effect of synergy
In Section \ref{sec:harmonyclash}, we discussed how pairs of granularities can have empty combinations either due to structure of calendar, event location or duration or due to the way they are build. In this section, we will see how these empty combinations affect the visualization when a dependent variable is plotted against these granularities.

@@

Hence we redefine harmony and clashes as follows : harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. Hence, we should avoid plotting clashes as they hinder the exploratory process by having missing combinations of time granularities in each panel.

@@

## Effect of number of observation

@@

Even with harmonies, problems might occur due to rarely occuring categories.

 Suppose we have $T$ hourly observations, and $C_1$ with $n_1$ categories and $C_2$ with $n_1$ categories. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For emperical estimates, even $k=10$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

@@

Let us take a specific example to illustrate this. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

### Unevenly distributed events

Even when there are no rarely occuring events, when number of observations are not evenly distributed between different categories of the time granularities.Large differences between number of observations between and across facets, might have an impact on the estimation of dfistribution. We use Gini's measure to compute if differences in number of observations across or within facets are significantly different.

@@

<!-- <!-- Existing approaches, why we need to drill down, or why we want to -- probability distributions --> -->
<!-- It is common to see aggregates of usage across households, total kwh used each half-hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.  -->

@@

In Figure \ref{fig:day_fortnight} and \ref{fig:day_fortnight_agg} , the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy beahvior of these two customers. In Figure \ref{fig:day_fortnight_agg} it can be seen that for most days of the fortnight, the second household has much less consumption than the first one. However, Figure \ref{fig:day_fortnight} adds to the beahvior by looking at the distribution. If we consider letter value F as a regular behavior and letter values beyond F as not-so-regular behavior, we can conclude that the regular behavior of the first household is more stable than the second household. However, the distribution of tail of the first household is more variable, observed through distinct letter values, implying that their not-so-regular behavior is quite extreme. This shows, how looking at the distribution of the dependent variable can throw more light on the energy behavior of the customers, which are lost using aggregate or summary statistics.

```{r day_fortnight_agg}
library(lvplot)
library(ggplot2)
library(dplyr)
library(tibble)
library(gravitas)
library(tidyr)
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704, 10017936)) %>% 
#   create_gran("day_fortnight") %>%
#   as_tibble() %>% 
#   group_by(customer_id, day_fortnight) %>% 
#   summarise(median= log(median(general_supply_kwh)),
#             total = log(sum(general_supply_kwh))) %>% 
#   gather("daily_consumption", "value", - c(customer_id, day_fortnight)) %>% 
#   ggplot2::ggplot(aes(
#     x = as.factor(day_fortnight),
#     y = value, 
#     color = daily_consumption,
#     group = daily_consumption)) +
#   geom_line() +
#   xlab("day_fortnight") +
#   ylab("general_supply_kwh") + 
#   facet_wrap(~customer_id) + 
#   theme_bw()

@@

While trying to explore the energy behavior of these customers systematically, the first thing we should have at our disposal for examining periodicities of energy behavior across time granularities is to know the number of time granularities we can look at exhaustively. If we consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, fortnight, month, quarter, semester, year), the following time granularities can be considered for this analysis.

@@

The interval of this tsibble is 30 minutes, and hence the default for search_gran in this case, provides temporal granularities ranging from half-hour to year. If these options are considered too many, the default options can be modified to limit the possibilities. For example, the most coarse temporal unit can be set to be a “month”.

@@

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
```

@@

- if the regular behavior is regular and extreme behavior extreme or vice versa
- if the weekend and weekday behavior are different for most of these households.

Figure \ref{fig:wknd_wday50} shows the overlay quantile plot of energy consumption of 10 households. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile.  It can be observed that the median is very closer to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. Weekend and weekday behavior is not very different for these ten customers, implying the first customer is more typical than the second customer when compared to these ten customers.

```{r wknd_wday50}
smart_meter10 %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = FALSE, 
            quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt()
```

```{r wknd_wday50_df}
smart_meter10 %>% 
  prob_plot("month_year",
                            "day_fortnight",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = TRUE, 
            quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt(breaks = c(0.1, 0.25, 0.5, 1:3))
```

@@

library(ggplot2)
smart_meter10 %>%
  prob_plot("month_year", "hour_day",
    response = "general_supply_kwh",
    plot_type = "quantile",
    quantile_prob = c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99),
    symmetric = TRUE) +
  ggtitle("") 
```

>>>>

The grammar of graphics introduced a framework to incorporate underlying features for statistical graphics to construct them by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is a modification over the former, allows us to produce graphics using the same structured thinking that is used to design analysis. Briefly, any graphics are made up of **Layers**, **Facet**,**Scale** and a  **coordinate system**. **Layers** are made up of **data**, **mapping**, **statistical transformation**, **geometric objects** and **position adjustments**.

@@

There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or empirical statistical methods. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. Density plots which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilties in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. The density based visualizations are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

@@

Given a context, it is good to be conversant with the benefits and challenges while choosing a distribution plot. As a general rule, if we have too many categories, the quantile plots are useful for camparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior.

### Effect of levels of temporal granularities

The levels of the two granularities plotted have an impact on the choice of plots since space and resolution might become a problem if the number of levels are too high. The criteria for different levels could be based on usual cognitive power while comparing across facets and display size available to us. Plot choices will also vary depending on which granularity is placed on the x-axis and which one across facets.

@@

  - If levels of both granularity plotted are low/medium, then any distribution plots might be chosen depending on which feature of the distribution needs focus.

@@

  - If levels of any granularity plotted are more than medium, empirical based methods of distribution visualizing are preferred as they use less space by design than most density based methods.

@@

## Effect of synergy of time granularities

In Section \ref{sec:harmonyclash}, we discussed how pairs of granularities can have empty combinations either due to structure of calendar, event location or duration or due to the way they are build. In this section, we will see how these empty combinations affect the visualization when a dependent variable is plotted against these granularities.

@@

We redefine harmony and clashes as follows: harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. As a result, we should avoid plotting clashes as they hinder the exploratory process by having missing combinations of time granularities in each panel.

@@

## Effect of number of observations

Even with harmonies, visualizing probability distributions can be misleading either due to rarely occuring categories or unevenly distributed events.

@@

Suppose we have $T$ observations, and two cyclic granularities -  $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge mnk$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For a decile plot, even $k=10$ may be acceptable, but for density estimates, we would need $k \ge 30$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

### Unevenly distributed events (TODO)

Even when there are no rarely occuring events, number of observations might vary hugely across within or across each panel. This might happen due to missing observations in the data or uneven locations of events in time domain. We use Gini's measure to compute if differences in number of observations across or within facets are significantly different.

@@

<!-- <!-- Existing approaches, why we need to drill down, or why we want to -- probability distributions --> 
<!-- It is common to see aggregates of usage across households, total kwh used each half-hour by state, for example, because energy companies need to understand maximum loads that they will have to plan ahead to accommodate. But studying overall energy use hides the distributions of usage at finer scales, and making it more difficult to find solutions to improve energy efficiency.  -->

@@

<!-- In Figure \ref{fig:day-fortnight} and \ref{fig:day-fortnight-agg} , the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy beahvior of these two customers. -->

In Figure \ref{fig:day_fortnight}, the smart meter data is filtered for two customers to illustrate what kinds of insights can be drawn for the energy beahvior of these two customers. it can be seen that for most days of the fortnight, the second household has much less consumption than the first one. However, there is additional information which we can derive looking at the distribution. If we consider letter value F as a regular behavior and letter values beyond F as not-so-regular behavior, we can conclude that the regular behavior of the first household is more stable than the second household. However, the distribution of tail of the first household is more variable, observed through distinct letter values, implying that their not-so-regular behavior is quite extreme. This shows, how looking at the distribution of the dependent variable can throw more light on the energy behavior of the customers, which are lost using aggregate or summary statistics.

```{r day-fortnight_agg, echo = FALSE}
library(lvplot)
library(ggplot2)
library(dplyr)
library(tibble)
library(gravitas)
library(tidyr)
# smart_meter10 %>% 
#   filter(customer_id %in% c(10006704, 10017936)) %>% 
#   create_gran("day_fortnight") %>%
#   as_tibble() %>% 
#   group_by(customer_id, day_fortnight) %>% 
#   summarise(median= log(median(general_supply_kwh)),
#             total = log(sum(general_supply_kwh))) %>% 
#   gather("daily_consumption", "value", - c(customer_id, day_fortnight)) %>% 
#   ggplot2::ggplot(aes(
#     x = as.factor(day_fortnight),
#     y = value, 
#     color = daily_consumption,
#     group = daily_consumption)) +
#   geom_line() +
#   xlab("day_fortnight") +
#   ylab("general_supply_kwh") + 
#   facet_wrap(~customer_id) + 
#   theme_bw()

@@

While trying to explore the energy behavior of these customers systematically across time granularities, the first thing we should have at our disposal is to know the number of time granularities we can look at exhaustively. If we consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, fortnight, month, quarter, semester, year), the following time granularities can be considered for this analysis.

@@

The interval of this tsibble is 30 minutes, and hence the default in this case, provides temporal granularities ranging from half-hour to year. If these options are considered too many, the default options can be modified to limit the possibilities. For example, the most coarse temporal unit can be set to be a “month”.

@@

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)

smart_meter10 %>% gran_advice("wknd_wday", "hour_day")

@@

- if energy distribution is skewed towards extreme?

- if the weekend and weekday behavior are different for most of these households?

Figure \autoref{fig:wknd-day50} shows the quantile plot of energy consumption of 50 households. Similar quantiles for weekend and weekdays indicate that behaviors of most of these customers do not alter between weekdays and weekends.

Figure \autoref{fig:my-hd} shows the area quantile plots of energy consumption across hours of the day faceted by months of the year. The black line is the median, whereas the pink band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile.  It can be observed that the median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. Moreover, only the pink band changes significantly in winter months (May - August). The median consumption or width of other bands doesn't vary much across seasons, implying extreme behavior or extreme customers does not vary across seasons much. Most of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that offcourse the level of quantiles increased too in winter months, the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.

```{r wknd-wday50}
smart_meter <- 
  read_rds("data/sm_cust50.rds")

smart_meter %>% 
  prob_plot("wknd_wday",
                            "hour_day",
                            response = "general_supply_kwh",
                            plot_type = "quantile",
            symmetric = FALSE, 
            quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) +
  scale_y_sqrt()
```

<!-- # ```{r wknd-wday50_df} -->
<!-- # smart_meter10 %>%  -->
<!-- #   prob_plot("month_year", -->
<!-- #                             "day_fortnight", -->
<!-- #                             response = "general_supply_kwh", -->
<!-- #                             plot_type = "quantile", -->
<!-- #             symmetric = TRUE,  -->
<!-- #             quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) + -->
<!-- #   scale_y_sqrt(breaks = c(0.1, 0.25, 0.5, 1:3)) -->
<!-- # ``` -->

@@

library(ggplot2)
smart_meter %>%
  prob_plot("month_year", "hour_day",
    response = "general_supply_kwh",
    plot_type = "quantile",
    quantile_prob = c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99),
    symmetric = TRUE) +
  ggtitle("") + scale_y_sqrt(breaks = c(0.01, 0.1, 0.5, 1:3))
```
